The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-08-05 paper/iphas-dr2-rev1.tex
Included edit by MJI

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archiving.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ {\bf extragalactic} survey programme carried out 
between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

{ \bf
Object detection and parametrisation is then carried out
using the standard methods in use by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated
by computing the sigma-clipped median intensity
in a grid of 64$\times$64 pixel bins across the image,
which is robust against contaminating sources.
Bilinear interpolation is then used to obtain an estimate 
of the background level at each pixel.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring pixels
in which the intensity exceeds the background level by at least $1.25\sigma$.
\item The object are parametrised:
positions are obtained by computing 
the intensity-weighted isophotal centre-of-gravity
of each object,
and photometry is obtained 
by measuring the intensity
in a series of soft-edged circular apertures of increasing diameter 
(1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6 and 6\farcs 6).
Objects are also classified morphologically
-- stellar, extended or noise --
based on their intensity as a function of aperture size.
\item The parametrisation of overlapping objects
is refined by simultaneous fitting of
soft-edged apertures to each blend,
effectively carrying out profile fitting where it is necessary.
We note that the parametrisation of these blended objects is naturally
less reliable than that of single, unconfused sources,
which is why they are flagged in the catalogue
(to be explained in \S\ref{sec:catalogue}).
\end{enumerate}

Having carried out object detection,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
using a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the field distortion of the camera.
The parameters of this solution
are then progressively refined by fitting against the
Two-Micron All Sky Survey \citep[2MASS; ][]{Skrutskie2006}.
We did not include proper motion information during this step,
because each CCD typically contains
hundreds of 2MASS reference stars.
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector
\citep{Gonzalez-Solares2008}.

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the position offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.
We note that both UCAC4 and 2MASS -- and hence also IPHAS --
are calibrated in the International Celestial Reference System (ICRS).
%We note that the mean residual is  
%is $-0.006 \pm 0.086$ in R.A.
%and $0.012\pm0.073$ in Dec.

@@

@@

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
{ \bf As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence
for degraded photometry up to ellipticities of 0.3.
We have excluded a small number of exposures
which exceeded $e = 0.3$.}

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(set in the pipeline at 0.09 for $r$ and 0.05 for $i$ as representative
averages for the telescope site),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used
(we adopt a 2\farcs 3-diameter circular aperture by default).
{\bf
Finally, \textsc{percorr} is a term used to correct 
for the small difference in internal gain
computed using the twilight flats
compared to the much redder typical astronomical objects. 
It is estimated by making a robust average of the dark sky levels 
measured on each detector during an observing run
(the correction is $0.01\pm0.01$ on average in $i$ and
$0.00\pm0.00$ in $r$ and \ha).
All these quantities correspond to header keywords
in the FITS files produced by the CASU pipeline. }

>>>>

All raw IPHAS data were transferred to the Cambridge Astronomical Survey Unit 
(CASU) for initial processing and archiving.  The procedures used by CASU 
were originally devised for the INT Wide Field imaging Survey 
\citep[WFS;][]{McMahon2001,Irwin2005}, which was a 200 deg$^2$ 
{\bf extragalactic} survey programme carried out between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination, we have been 
able to benefit from the existing WFS pipeline. A description of the 
processing steps can be found in \citet{Irwin2001}. Its application to IPHAS 
has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and some elements of the the source code are available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the imaging processing part of the pipeline takes care of bias 
subtraction, the linearity correction, flat-fielding with internal gain
correction, and de-fringing for the i-band.

{ \bf
Object detection and parametrisation is then carried out
using the standard methods developed by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated by first computing an iteratively
sigma-clipped median intensity on a grid of 64$\times$64 pixel bins across 
the image from each detector.  This is usually robust against contaminating 
sources corrupting the background level. The resulting array of background
values is then filtered to further reduce the effect of large objects on
the local background level. Bilinear interpolation is then used to obtain an 
estimate of the background level at the orginal image pixel scale.
\item To improve faint object detection, each image is smoothed with a matched
detection filter which is used in conjunction with the unsmoothed image
for object detection and parameterisation.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring contiguous pixels
in which the intensity exceeds the background level by at least $1.25\sigma$
on the matched filtered image. Objects are deblended using a sequence of
successively higher detection levels.
\item The objects are parametrised using the unsmoothed image at these pixel 
locations : positions are obtained based on an intensity-weighted isophotal 
centre-of-gravity of each object; whilst photometry is derived by measuring 
the intensity in a series of soft-edged circular apertures covering a range 
of diameters (1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6, 6\farcs 6 and 
9\farcs 2).  Objects are also classified morphologically
-- stellar, extended or noise -- based on their intensity as a function of 
aperture size and on their intensity-weighted second moments, where the latter
are used to derive an estimate of object ellipticity.
\item The parametrisation of overlapping objects is refined by simultaneous 
fitting of soft-edged apertures to each blend,
effectively carrying out ``top hat'' profile fitting where it is necessary.
We note that the parametrisation of these blended objects is naturally
less reliable than that of single, unconfused sources,
which is why they are flagged in the catalogue
(to be explained in \S\ref{sec:catalogue}).
\end{enumerate}

Having carried out object detection,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
using a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the (fixed) field distortion of the camera.
The parameters of this solution
are then progressively refined by fitting against the
Two-Micron All Sky Survey \citep[2MASS; ][]{Skrutskie2006}.
We did not include proper motion information during this step,
because each CCD typically contains
hundreds of 2MASS reference stars.
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector array
\citep{Gonzalez-Solares2008}.

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the position offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.
%We note that both UCAC4 and 2MASS -- and hence also IPHAS --
%not 100% true due to bulk pm of calibrators - Mike
%are calibrated in the International Celestial Reference System (ICRS).
%We note that the mean residual is  
%is $-0.006 \pm 0.086$ in R.A.
%and $0.012\pm0.073$ in Dec.

@@

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect noise-free imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions, tracking errors and photon plus readout noise.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
{ \bf As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence
for degraded photometry up to ellipticities of 0.3.
We have excluded a small number of exposures
which exceeded $e = 0.3$.}

@@

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(set in the pipeline at 0.09 for $r$ and 0.05 for $i$ as representative
averages for the telescope site),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used
(we adopt a 2\farcs 3-diameter circular aperture by default).
{\bf
Finally, \textsc{percorr} is a term used to correct 
for the small difference in internal gain
computed using the relatively blue twilight flats
compared to the much redder typical astronomical objects. 
It is estimated by making a robust average of the dark sky levels 
measured on each detector during an observing run
(the correction is $0.01\pm0.01$ on average in $i$ and
$0.00\pm0.00$ in $r$ and \ha).
All these quantities correspond to header keywords
in the FITS files produced by the CASU pipeline. }
